## Loading relevant libraries

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(skimr)
library(viridis)
```

## Loading the data

```{r, warning=FALSE, message=FALSE}
data <- read_delim("/Users/isaachasbani/Desktop/LBS/Data Visualization/am10/Group project/data.csv", delim = "#")
```

```{r}
summary(data)
```

## Check for missing values

```{r, warning=FALSE, message=FALSE}
missing_values <- data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>% 
  pivot_longer(everything(), names_to = "column", values_to = "missing_count")

missing_values
```

## Feature Engineering

```{r, warning=FALSE, message=FALSE}
#remove missing values
data <- data %>%
  filter(complete.cases(.))

# Feature engineering
data <- data %>% 
  # Rename "Track Name" column to "Track" for simplicity
  rename("Track" = "Track Name") %>% 
  mutate(
    # Extract the month from the "Date" column, with abbreviated month names (e.g., "Jan", "Feb")
    Month = month(Date, label = TRUE, abbr = TRUE),
    
    # Extract the day of the month from the "Date" column
    Day = day(Date),
    
    # Extract the year from the "Date" column
    Year = year(Date),
    
    # Create a binary variable "Featuring" to indicate if the track has a featured artist
    # It checks if the track name contains "feat." or "Feat." using `str_detect()`
    Featuring = ifelse(str_detect(Track, "feat.|Feat."), 1, 0)
  )
```

## Get an idea of the how the the streams of the top 200 songs are increasing over time

```{r, warning=FALSE, message=FALSE}
daily_streams <- data %>%
  
  # Filter out 2021 as we only have incomplete data (Jan to Jul)
  filter(Year < 2021) %>% 
  
  # Group the data by 'Date' to get daily streams
  group_by(Date) %>% 
  
  # Calculate total streams per day
  summarize(total_streams = sum(Streams)) %>%
  
  ggplot(aes(x = Date, y = total_streams)) +
  geom_line(color = "#1DB954", size = 1) +  
  
  # Add a smoothed line to show the trend
  geom_smooth(
    
    # Use LOESS method for smoothing (locally weighted regression)
    method = "loess", 
    color = "#191414",
    
    # Disable confidence interval shading around the smooth line
    size = 1, 
    se = FALSE)+
  labs(
  title = "Total streams of the Top-200 are increasing over time",
  subtitle = "Period: 2017-2021",
  caption = "Source: Spotify"
  ) +
  theme_minimal() +
  
  # Customize the appearance of axis titles and text with Spotify colors
  theme(
  axis.title = element_blank(),
  axis.text.x = element_text(face = "bold", size = 10, color = "#191414"),
  axis.text.y = element_text(face = "bold", size = 10, color = "#191414"),
  plot.title = element_text(size = 16, face = "bold", color = "#1DB954"),
  plot.subtitle = element_text(size = 12, face = "bold", color = "#191414"),
  plot.caption = element_text(size = 12, color = "#1DB954", hjust = 1, face = "bold"),
  
  )+
  scale_y_continuous(labels = function(x) sprintf("%.2fM", x / 1e6))

daily_streams
```

## Is this increase consistent for all genres?

```{r, warning=FALSE, message=FALSE}
data <- data %>%
  
  # Data Preprocessing
  
  # Step 1: Clean the Genre column to remove unwanted characters and extract the first genre from the list
  
  mutate(Genre = sub("^\\['([^']+)'.*$", "\\1", Genre)) %>%
  
  # Step 2: Create a new column 'genre_group' to categorize genres into broader groups
  mutate(
      genre_group = case_when(
      
      # Identify key words and categorize genres into Electronic, Pop, Hip Hop, Rock, Classical, and Latin
      str_detect(Genre, "dance|edm|house") ~ "Electronic",
      str_detect(Genre, "pop") ~ "Pop",
      str_detect(Genre, "rap|hip hop|trap") ~ "Hip Hop",
      str_detect(Genre, "rock") ~ "Rock",
      str_detect(Genre, "classical|orchestral") ~ "Classical",
      str_detect(Genre, "latin|reggaeton") ~ "Latin",
      
      # For genres not covered in specific groups
      TRUE ~ "Other"
    )
  ) %>%

  # Step 3: Remove rows from 2021, as we only have incomplete data
  filter(Year != 2021) 

# Step 4: CCreate monthly trends for each genre group
monthly_genre_trends <- data %>%
  
  # Round each date to the start of the month 
  mutate(Month_floor = floor_date(Date, unit = "month")) %>% 
  
  # Group by the rounded month (Month_floor) and genre group
  group_by(Month_floor, genre_group) %>% 
  
  # Total streams per month
  summarise(total_streams = sum(Streams)) %>%
  ungroup() %>%
  
  # Step 5: Calculate the proportion of total streams for each genre group in each month
  group_by(Month_floor) %>%
  
  # Calculate proportions
  mutate(proportion_streams = total_streams / sum(total_streams, na.rm = TRUE)) %>% 
  ungroup()

# Step 6: Calculate the initial proportion for each genre group at the start of the period
initial_proportions <- monthly_genre_trends %>%
  group_by(genre_group) %>%
  
  # Find the earliest month for each genre
  filter(Month_floor == min(Month_floor)) %>% 
  
  # Average if there are duplicates
  summarise(initial_proportion = mean(proportion_streams, na.rm = TRUE)) %>% 
  
  # Sort by initial proportion in descending order
  arrange(desc(initial_proportion)) 

# Step 7: Reorder genre groups based on their initial proportion to maintain a consistent color scheme
monthly_genre_trends <- monthly_genre_trends %>%
  mutate(
    genre_group = factor(genre_group, levels = initial_proportions$genre_group)
  )

ggplot(monthly_genre_trends, aes(x = Month_floor, y = proportion_streams, fill = genre_group)) +
  
  # Use an area chart to visualize the proportions of streams for each genre over time
  geom_area(alpha = 0.7) +  # White borders for clarity
  labs(
    title = "Pop dominates the music scene in the late 2010s",
    x = "",
    y = "Proportion of Total Streams",
    caption = "Source: Spotify"
  ) +
  
  # Percent y-axis and remove gaps
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  
  # Remove gaps on x-axis
  scale_x_date(expand = c(0, 0)) +
  
   # Use a clear color palette for genres
  scale_fill_brewer(palette = "Set3") +
  
  # Use a classic, clean theme
  theme_classic() +
  
  # Customize the appearance of axis titles and text with Spotify colors
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold",, color = "#1DB954"),
    plot.caption = element_text(size = 10, face = "bold", hjust = 1, color = "#1DB954"), # Green caption
    legend.position = "right",
    legend.title = element_blank()
  )
```

## How do the listening change over the year?

```{r, warning=FALSE, message=FALSE}
# Heat Map


# We create a data frame with information about the tiles we want to highlight

highlight_dates <- data.frame(
  
  # Dates for Christmas eve and day
  date_of_month = c(24, 25),
  
  # Dates are in December
  month_categorical = factor("Dec")
)


# Data Transformation:
data %>% 
  
  # Filter out 2021 due to incomplete data
  filter(Year != 2021) %>% 
  
  # We want our calculations to be on a daily basis for each month
  group_by(Month, Day) %>%
  
  # Calculate the total streams for each day of every month
  summarise(total_streams = sum(Streams)) %>%
  

# Plotting
  
  ggplot() + 
  
  aes(x = Day, 
      y = Month, 
      fill = total_streams) +
  
  # Heatmap with white borders
  geom_tile(color = "white") +
  
  # Viridis color scale (continuous) with reversed direction for better highlights
  scale_fill_viridis_c(
    name = "Total Streams",
    option = "viridis",
    direction = -1,
    labels = scales::label_comma(),
    guide = guide_colorbar(
      barwidth = 20, barheight = 0.3, position = "bottom"
    )
  ) +
  
  # Adding titles and caption
  labs(
    title = "Spotify Streaming Peaks Around Christmas",
    subtitle = "Daily streaming of top 200 Spotify hits (2017 - 2020)",
    caption = "Source: Spotify"
  ) +
  
  # Ensure x-axis displays all days in the month
  scale_x_continuous( breaks = 1:31) +
  
  # Minimal theme for a clean plot
  theme_minimal(base_family = "Lato") +
  
  # Customize plot elements for Spotify colors
  theme(
    panel.grid = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 7, color = "#191414"),
    axis.text.y = element_text(size = 8, color = "#191414"),
    legend.title = element_text(size = 10, hjust = 0.5, color = "#191414", face = "bold"),
    legend.title.position = "bottom",
    legend.text = element_text(size = 5, color = "#191414", hjust = 0.5),
    plot.caption = element_text(size = 8, color = "#1DB954", hjust = 1, face = "bold"),
    plot.title = element_text(size = 10, face = "bold", color = "#1DB954"),
    plot.subtitle = element_text(size = 9, color = "#191414")
  ) +
  
  # Highlight tiles for Dec 24 and Dec 25
  geom_tile(
    data = highlight_dates,
    aes(x = date_of_month, y = month_categorical),
    color = "red", fill = NA, size = 1.1)
```

## December trends

```{r, warning=FALSE, message=FALSE}
# Define tracks to highlight
tracks_to_highlight <- c("Dance Monkey", "Blinding Lights")

# Prepare and plot the data
christmas_data <- data %>% 
  # Filter the data to include only December (Month == "Dec")
  filter(Month == "Dec") %>% 
  
  # Group the data by 'Month' and 'Track' to calculate total streams for each track in December
  group_by(Month, Track) %>% 
  
  # Summarize the data to get the total streams for each track
  summarise(total_streams = sum(Streams), .groups = 'drop') %>% 
  
  # Arrange the data in descending order of total streams
  arrange(desc(total_streams)) %>% 
  
  # Select the top 10 tracks based on total streams
  slice_head(n = 10) %>% 
  
  # Reorder the 'Track' factor by 'total_streams' to ensure the bars are ordered in descending order
  mutate(
    Track = fct_reorder(Track, total_streams, .desc = TRUE),
    
    # Create a new column 'Category' to classify tracks into 'Christmas Themed Song' or 'Other Song'
    # If the track is in the 'tracks_to_highlight' list, it will be labeled as "Other Song"
    Category = if_else(Track %in% tracks_to_highlight, "Other Song", "Christmas Themed Song")
  ) %>% 
  
  # Create the plot
  ggplot(aes(x = Track, y = total_streams, fill = Category)) +
  
  # Create a bar plot with columns
  geom_col() +
  
  # Customize the fill color of the bars based on the 'Category' variable
  scale_fill_manual(
    name = "Song Category",
    values = c("Christmas Themed Song" = "#c60f0f", "Other Song" = "#1DB954")  # Red for Christmas songs, green for others
  ) + 
  
  # Apply a minimal theme (no gridlines, simple style)
  theme_minimal() +
  
  # Add labels, title, and caption to the plot
  labs(
    title = "8 of the Top 10 Songs in December are Christmas-Themed",  # Title of the plot
    y = "Total Streams",  # Label for the y-axis
    caption = "Source: Spotify"  # Data source caption
  ) +
  
  # Customize the appearance of the plot for Spotify colors
  theme(
    panel.grid = element_blank(),  # Remove gridlines
    axis.text.x = element_text(
      angle = 45, hjust = 1, vjust = 1, 
      face = "bold", size = 14, color = "#191414" 
    ),
    axis.text.y = element_text(
      face = "bold", size = 12, color = "#191414" 
    ),
    axis.title = element_blank(),  # Remove axis titles (no need for them here)
    plot.title = element_text(
      size = 18, face = "bold", color = "#1DB954" 
    ),
    plot.caption = element_text(
      size = 12, color = "#1DB954", 
      hjust = 1, face = "bold"
    )
  ) +
  
  # Add labels on top of each bar displaying the total streams in millions
  geom_text(
    aes(label = sprintf("%.2fM", total_streams / 1e6)),  # Format the stream count in millions with 2 decimals
    vjust = -0.5, size = 5  # Position the labels slightly above the bars
  ) +
  
  # Format the y-axis labels to display numbers in millions with commas
  scale_y_continuous(labels = scales::comma_format(scale = 1e-6, suffix = "M"))

christmas_data  # Show the plot created
```

## Featuring vs Solo songs. Which ones perform better?

### Monthly data

```{r, warning=FALSE, message=FALSE}
# Calculate monthly stats with confidence intervals
monthly_ci <- data %>%
  
  # Group by month and whether the song features an artist (used our binary variable)
  group_by(Month, Featuring) %>%
  summarize(
    Average_Streams = mean(Streams),
    Lower_CI = mean(Streams) - 1.96 * (sd(Streams) / sqrt(n())),
    Upper_CI = mean(Streams) + 1.96 * (sd(Streams) / sqrt(n())),
    .groups = 'drop'
  )

# Plot with shaded confidence intervals
# Correct labeling and consistent mapping for Featuring
monthly_ci$Featuring <- factor(monthly_ci$Featuring, levels = c(1, 0), labels = c("Featuring", "Solo"))

# Plot with corrected labels and white background
# Plot with no legend title
monthly_ci_plot <- ggplot(monthly_ci, aes(x = Month, y = Average_Streams, color = Featuring, group = Featuring)) +
  
  # Shaded confidence area
  geom_ribbon(aes(ymin = Lower_CI, ymax = Upper_CI, fill = Featuring), alpha = 0.2, color = NA) + 
  
  # Line graph
  geom_line(size = 1.5, alpha = 0.8) + 
  
  # Stylish points
  geom_point(size = 4, shape = 21, fill = "white", stroke = 1.5) + 
  
  # Customize the color scheme for Featuring and Solo
  scale_color_manual(values = c("Featuring" = "#1DB954", "Solo" = "#191414"), name = NULL) + # Remove legend title
  scale_fill_manual(values = c("Featuring" = "#1DB954", "Solo" = "#191414"), name = NULL) + # Remove legend title
  labs(
    title = "Featurings Heat Up Summer, Solo Songs Chill Christmas",
    caption = "Source: Spotify"
  ) +
  theme_minimal() +
  
  # Customize the appearance of the plot elements
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12, color = "#191414", face = "bold"),
    axis.text.y = element_text(size = 12, color = "#191414", face = "bold"),
    axis.title = element_blank(), # Removes axis titles
    plot.title = element_text(size = 14, face = "bold", color = "#1DB954"),
    panel.grid.major = element_line(color = "#D3D3D3", size = 0.5),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#FFFFFF", color = NA),
    plot.background = element_rect(fill = "#FFFFFF", color = NA),
    legend.background = element_rect(fill = "#FFFFFF"),
    legend.key = element_rect(fill = "#FFFFFF"),
    legend.text = element_text(color = "#191414", size = 12),
    legend.position = "bottom",
    plot.caption = element_text(size = 12, color = "#1DB954", hjust = 1, face = "bold")
  )+
  
  # Format the y axis to display stream numbers in millions
  scale_y_continuous(labels = function(x) sprintf("%.2fM", x / 1e6))

monthly_ci_plot
```

### Yearly data

```{r, warning=FALSE, message=FALSE}
yearly_ci <- data %>%
  
  # Detect 'feat.' or 'Feat.' in track names and label accordingly
  mutate(Featuring = ifelse(str_detect(Track, "feat.|Feat."), 1, 0)) %>% # Adjusted column name
  
  # Group the data by Year and whether the song is Featuring or Solo
  group_by(Year = year(Date), Featuring) %>%
  
  # Summarize the data by calculating average streams and confidence intervals for each year and category
  summarize(
    Average_Streams = mean(Streams),
    Lower_CI = mean(Streams) - 1.96 * (sd(Streams) / sqrt(n())),
    Upper_CI = mean(Streams) + 1.96 * (sd(Streams) / sqrt(n())),
    .groups = 'drop'
  )

# Adjust the 'Featuring' column to be a factor with custom labels for better readability in the plot
yearly_ci$Featuring <- factor(yearly_ci$Featuring, levels = c(1, 0), labels = c("Featuring", "Solo"))

# Create the yearly plot
yearly_ci_plot <- ggplot(yearly_ci, aes(x = Year, y = Average_Streams, color = Featuring, group = Featuring)) +
  
  # Shaded confidence area
  geom_ribbon(aes(ymin = Lower_CI, ymax = Upper_CI, fill = Featuring), alpha = 0.2, color = NA) + 
  
  # Line graph
  geom_line(size = 1.5, alpha = 0.8) + 
  
  # Stylish points
  geom_point(size = 4, shape = 21, fill = "white", stroke = 1.5) + 
  
  # Customize the color scheme for Featuring and Solo
  scale_color_manual(values = c("Featuring" = "#1DB954", "Solo" = "#191414")) +
  scale_fill_manual(values = c("Featuring" = "#1DB954", "Solo" = "#191414")) +
  labs(
    title = "Featurings have over performed solo songs after COVID-19",
    caption = "Source: Spotify",
    color = NULL,    # Removes the color legend title
    fill = NULL      # Removes the fill legend title
  ) +
  theme_minimal() +
  
  # Customize for Spotify colors
  theme(
    axis.text.x = element_text(size = 12, color = "#191414", face = "bold"),
    axis.text.y = element_text(size = 12, color = "#191414", face = "bold"),
    axis.title = element_blank(), # Removes axis titles
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold", color = "#1DB954"),
    panel.grid.major = element_line(color = "#D3D3D3", size = 0.5),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#FFFFFF", color = NA),
    plot.background = element_rect(fill = "#FFFFFF", color = NA),
    legend.background = element_rect(fill = "#FFFFFF"),
    legend.key = element_rect(fill = "#FFFFFF"),
    legend.text = element_text(color = "#191414", size = 12),
    # Remove the legend titles by setting them to NULL
    legend.title = element_blank(),
    legend.position = "bottom",
    plot.caption = element_text(size = 12, color = "#1DB954", hjust = 1, face = "bold")
  ) +
  scale_y_continuous(labels = function(x) sprintf("%.2fM", x / 1e6))

yearly_ci_plot

```

### Who is the top performing artist during the period?

```{r, warning=FALSE, message=FALSE}
# Sum total streams per artist and get the top 10
top_artists <- data %>%
  group_by(Artist) %>%
  summarize(total_streams = sum(Streams)) %>%
  top_n(10, total_streams)

# Plot with Spotify colors

# Convert streams to billions
ggplot(top_artists, aes(x = reorder(Artist, total_streams), y = total_streams / 1e9)) + 
  
  # Spotify green color
  geom_bar(stat = "identity", fill = "#1DB954") + 
  
  # Add 'B' for billions
  geom_text(aes(label = paste0(scales::comma(total_streams / 1e9, accuracy = 0.01), "B")),
            # Bold labels
            hjust = -0.2, size = 3.5, color = "black", fontface = "bold") + 
  coord_flip() +
  labs(
    title = "Post Malone was the most streamed artist from 2017 to 2021",
    
    # Remove the x-axis caption
    x = "", 
    
    # Remove the y-axis label
    y = "", 
    
    # Add source caption
    caption = "Source: Spotify" 
  ) +
  
  # Use Arial font
  theme_minimal(base_family = "Arial") +
  theme(
    plot.background = element_blank(),      # White background
    panel.background = element_blank(),     # White panel
    text = element_text(color = "black"),                # Black text
    axis.text.x = element_blank(),          # Remove x-axis text
    axis.ticks.x = element_blank(),         # Remove x-axis ticks
    axis.line.x = element_blank(),          # Remove x-axis line
    axis.title.y = element_blank(),         # Remove y-axis title
    axis.title.x = element_blank(),         # Remove x-axis title
    axis.text.y = element_text(face = "bold"), # Make y-axis text (artists) bold
    plot.title = element_text(size = 16, face = "bold", color = "#1DB954"),  # Spotify green for title
     plot.caption = element_text(size = 12, color = "#1DB954", hjust = 1, face = "bold"), # Green caption
    legend.position = "none"                             # Remove legend if any
  )
```

### What causes Post Malone's streams to increase?

```{r, warning=FALSE, message=FALSE}

# These are the dates when Post Malone's albums were released:
# Dates of "Beerbongs & Bentleys" and "Hollywood's Bleeding"
highlight_dates <- as.Date(c("2018-04-20", "2019-08-27"))

# Filter data for Post Malone and calculate daily song counts in the Top 200
post_malone <- data %>% 
  filter(Artist == "Post Malone") %>% 
  group_by(Date) %>% 
  summarise(count = n())

# Create a new dataframe with the highlight dates and labels for the album releases
highlight_counts <- post_malone %>%
  filter(Date %in% highlight_dates) %>% 
  mutate(label = c("Album Release:\nBeerbongs & Bentleys", "Album Release:\nHollywood's bleeding"))

# Create the plot to visualize Post Malone's song counts and highlight the album release dates
post_plot <- ggplot(post_malone, aes(x = Date, y = count)) +
  
  # Add a line graph to show the trend of the number of Post Malone songs in the Top 200 over time
  geom_line(size = 1, color = "#1DB954") +
  
  # Highlight the specific dates (album releases) with red points
  geom_point(
    data = highlight_counts, 
    aes(x = Date, y = count), 
    shape = 21,          
    color = "red",       
    fill = NA,      
    size = 8,             
    stroke = 0.8         
  ) +
  
  # Add text labels next to the points to indicate the album releases
  geom_text(
    data = highlight_counts,
    aes(x = Date, y = count, label = label),
    vjust = 2,            
    hjust = 0,           
    color = "red",
    size = 3,
    fontface = "bold") + 
  labs(
    title = "Songs in the Top 200 Spike After an Album Release",
    subtitle = "Number of Post Malone's Songs in the Top 200 (2017-2021)", 
    caption = "Source: Spotify"
  ) +
  theme_minimal() +
  
  # Customize axis labels, title, and gridlines for Spotify colors
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(face = "bold", size = 10, color = "#191414"),
    axis.text.y = element_text(face = "bold", size = 10, color = "#191414"),
    plot.title = element_text(size = 16, face = "bold", color = "#1DB954"),
    plot.subtitle = element_text(size = 12, face = "bold", color = "#191414"),
    plot.caption = element_text(size = 12, color = "#1DB954", hjust = 1, face = "bold")
  ) +
  scale_x_date(
    date_labels = "%b\n%Y",
    date_breaks = "4 months" 
  ) 


post_plot
```

### Save the cleaned data in the directory

```{r, warning=FALSE, message=FALSE}
write.csv(data, "data_cleaned.csv", row.names = FALSE)
```
